---
title: "Lab 13 -143"
author: "Ethan"
date: "2/21/2023"
output: pdf_document
---
Files "GSE37704_metadata.csv" and "GSE37704_featurecounts.csv" are downloaded from the class website.

```{r}
# Will be using DESeq2. Call in pkg
library(DESeq2)
```

```{r}
# Load in files as variables
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata
colData = read.csv(metaFile, row.names=1)
head(colData)

# Import countdata
countData = read.csv(countFile, row.names=1)
head(countData)
```

Q1. Complete the code below to remove the troublesome first column from countData

```{r}
# Note we need to remove the odd first $length col
# I added '-1' to remove the first col
countData <- as.matrix(countData[,-1])
head(countData)
```


Q2. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns).

Tip: What will rowSums() of countData return and how could you use it in this context?

```{r}
# Filter count data where you have 0 read count across all samples.
noZero <- rowSums(countData)>0
countData = countData[noZero,]
head(countData)
nrow(countData)
```

# Running DESeq2

```{r}
# From the lab guide
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```
```{r}
# Look at the variable 'dds' output
dds

# Store data in 'res'
res <- results (dds)
```

Q3. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.

4349 genes are upregulated and 4396 genes are downregulated

```{r}
# Following instructions above
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
summary(res)
```
# Volcano Plot
```{r}
# Basic volc plot
plot( res$log2FoldChange, -log(res$padj) )
```

Q4. Improve this plot by completing the below code, which adds color and axis labels

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

# Gene Annotation

Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below.

New Installations:
BiocManager::install("AnnotationDbi")
BiocManager::install("org.Hs.eg.db")

```{r}
# Call in pkgs
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
# Use mapID() to add annotations. Answering Q above.
columns(org.Hs.eg.db)
res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```

Q. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory.

```{r}
# Order res by pvalue
res = res[order(res$pvalue),]
# Save res as new csv file
write.csv(res, file = "deseq_results.csv")
```

# Pathway Analysis
New Installation:
BiocManager::install( c("pathview", "gage", "gageData") )

```{r}
# Load packages for KEGG
library(pathview)
library(gage)
library(gageData)
```

```{r}
# Loads data sets
data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

```{r}
# Set new variable for fold change (from the DESeq analysis)
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)

attributes(keggres)
# Look at the first few down (less) pathways
head(keggres$less)
```

```{r}
# Pathview makes a pathway visual map.
# kegg.native=FALSE displays as a pdf graph
# hsa04110 = cell cycle
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```
# Focus on top 5 upregulated pathways here for demo purposes only

```{r}
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

```{r}
# Top 5 pathways upregulated
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-reguled pathways?

```{r}
# use keggres$less
keggrespathwaysdown <- rownames(keggres$less)[1:5]

# Extract the 8 character long IDs part of each string
keggresids2 = substr(keggrespathwaysdown, start=1, stop=8)
keggresids2
```

```{r}
# Top 5 pathways downregulated
pathview(gene.data=foldchanges, pathway.id=keggresids2, species="hsa")
```

# Gene Ontology

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

head(gobpres$less)
```

# Reactome Analysis
New Installation:
BiocManager::install("ReactomePA")

```{r}
library("ReactomePA")
```

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

Q: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

Top Pathways from Reactome: Endosomal/Vacuolar pathway and Cell Cycle
GO:0048285 Organelle Fission is a top hit from KEGG (and it's part of the cell cycle)
Maybe different considerations of defining a pathway could cause differences in either method.


