---
title: "Lab-17-MiniAssn"
author: "Ethan"
date: "3/7/2023"
output: html_document
---

## Class 17: COVID19 Vaccine Rate Mini Assignment

Data Import:
Download “Statewide COVID-19 Vaccines Administered by ZIP Code” from website
Import it into the Project folder and read it with the code below

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

Q1. What column details the total number of people fully vaccinated?
Column 10

Q2. What column details the Zip code tabulation area?
Column 2

Q3. What is the earliest date in this dataset?
2021-01-05
```{r}
# Use table to look at all the dates and find the earliest one. It's already in order
table(vax["as_of_date"])
# This works too
# vax[sort("as_of_date")]
```

Q4. What is the latest date in this dataset?
2023-02-28
```{r}
# Table is already in order by dates. We can just observe the last row
tail(vax)
vax$as_of_date[nrow(vax)]
```

```{r}
# Get a quick overview of dataset
skimr::skim(vax)
```

Q5. How many numeric columns are in this dataset?
13 columns. Under column type frequency: numeric:

Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?
16525 NA vales
```{r}
# Use 'is.na' to give a T/F matrix
# table() it to give you the count of each
table(is.na(vax$persons_fully_vaccinated))
```


Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?
8.29% of persons_fully_vaccinated values are missing
```{r}
# sum() adds up the number of TRUE. We can store this into n.missing to use it
n.missing <- sum(is.na(vax$persons_fully_vaccinated))
round((n.missing/ nrow(vax))*100, 2)
```
# Working with Dates
New Installation: (In console)
install.packages("lubridate")

```{r}
# Call in lubridate pkg
library("lubridate")
```

```{r}
# Test some of its functions
today()
```

Note: today() - vax$as_of_date[1] this won't work. You can't directly subtract dates.

```{r}
# Put into ymd() format first
today() - ymd("2021-01-05")

# Relating it to a working dataframe
vax$as_of_date <- ymd(vax$as_of_date)
today() - vax$as_of_date[1]
```

Q9. How many days have passed since the last update of the dataset?
7 days (as of 2023-03-07)

```{r}
# Find the most recent date in the data set
today() - ymd(vax$as_of_date[nrow(vax)])
```

Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?
113 unique dates
```{r}
# Use table and count the rows
nrow(table(vax$as_of_date))
```
# Working with ZIP codes
New Installations: (On console)
install.packages("zipcodeR")

```{r}
# Call in zipcodeR pkg
library("zipcodeR")
```

```{r}
# geocode_zip gives you the lat and lng for a certain zip code
geocode_zip('92037')
# Inputting 2 zip codes with zip_distance gives you the distance between them (IN MILES)
zip_distance('92037','92109')
```

```{r}
# reverse_zipcode pulls out all the related information tied to the zip code
# we can store this in zip_data
zip_data <- reverse_zipcode(c('92037', "92109"))
```

# Focus on the San Diego Area
2 main methods:
  1 using base R
  2 using dplyr pkg

```{r}
# Subset to San Diego county only areas: Method 1
sd <- vax$county == "San Diego"
sdx <- vax[sd,]
```

```{r}
# Call in dplyr pkg
library(dplyr)
```

```{r}
# Subset to San Diego county only areas: Method 2
sd.2 <- filter(vax, county == "San Diego")
nrow(sd.2)
```
Q11. How many distinct zip codes are listed for San Diego County?
107 distinct zip codes in SD county
```{r}
# nrow used on table of San Diego df (sdx)
# Note: sd2 and sdx are both "San Diego-sorted" dataframes
nrow(table(sdx$zip_code_tabulation_area))
```

Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?
92154 has the largest 12+ Population in this dataset
```{r}
# Find the index of the row with largest tot_population
# Use the index to find the zip code it matches with
high <- which.max(sdx$age12_plus_population)
sdx[high ,"zip_code_tabulation_area"]
```

Sort the San Diego counties by most recent date and answer the next 2 questions

```{r}
# Using dplyr to filter the df
sd.date <- filter(vax, county == "San Diego" & as_of_date == "2023-02-28")
```

Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2023-02-28”?
0.7400878 or 74.01%

```{r}
# remove the NA rows first & put new df for ease
sd.ppfv <- sd.date[is.na(sd.date$percent_of_population_fully_vaccinated) == 0, ]

# take average of new dataframe without the NA rows
mean(sd.ppfv$percent_of_population_fully_vaccinated)
```

Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-11-15”?

```{r}
# Using base R plots
hist(sd.ppfv$percent_of_population_fully_vaccinated, breaks=20)
```

# Focus on UCSD and La Jolla

```{r}
# Use dplyr pkg to filter by area code
# sd is the T/F dataframe & sdx is the actual dataframe that's sorted
ucsd <- filter(sdx, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
library(ggplot2)
```

```{r}
# Fill in the ggplot code from the lab manual
ucsd.plot <- ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
ucsd.plot
```

```{r}
# filter again matching population of 92037 on “2022-02-28” (most recent date)
vax.36 <- filter(vax, age5_plus_population > 36144 & as_of_date == "2023-02-28")
```

Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2023-02-28”. 
0.7213331 or 72.13%
```{r}
mean(vax.36$percent_of_population_fully_vaccinated)
```

Add this as a straight horizontal line to your plot from above with the geom_hline() function?
```{r}
# Adding onto existing ucsd.plot wtih geom_hline()
ucsd.plot + geom_hline(yintercept = 0.7213331, col="red", linetype = "dashed")
```

Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2023-02-28”?

```{r}
# Use fivenum() on data
# Output of min, lowerQ, median, upperQ, max
fivenum(vax.36$percent_of_population_fully_vaccinated)

# Individually Calculate Mean
vax.36_mean <- mean(vax.36$percent_of_population_fully_vaccinated)

sixnum <- c(fivenum(vax.36$percent_of_population_fully_vaccinated),vax.36_mean)

# or summary(vax.36percent_of_population_fully_vaccinated) gives all 6
```

Q18. Using ggplot generate a histogram of this data.
```{r}
hist <- ggplot(vax.36) +
  aes(x=percent_of_population_fully_vaccinated) +
  xlim(0,1) +
  geom_histogram() + geom_vline(aes(xintercept=vax.36_mean))
hist
```

Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
The mean of percent_of_population_fully_vaccinated of vax.36 is 0.7213331
Both averages 0.550296 and 0.694572 are below the average

```{r}
# Calc average for 92040
vax %>% filter(as_of_date == "2023-02-28") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)

# Calc average for 92109
vax %>% filter(as_of_date == "2023-02-28") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```


Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(0,1) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rates Across CA",
       subtitle="only areas with population above 36k are shown") +
  geom_hline(yintercept = vax.36_mean, linetype="dashed")
```




